---
title: 'Session 10: Sampling variability and Effect size'
date: "May 13, 2021"
output: html_document
---

## Agenda

* How are sampling distributions affected by sample size and variance?
* Intro to standardized effect sizes; Cohen's *d* 

## Scripts

The R markdown files from session 10 are available below.

```{r, include = FALSE, eval = FALSE}

# install and load these packages
devtools::install_github('yihui/xfun')

# embed_file() requires a few more packages
xfun::pkg_load2(c('base64enc', 'htmltools', 'mime'))
```

```{r include = TRUE, echo = FALSE}
# a embed single file
xfun::embed_files(c("session_10A.Rmd", 
                    "session_10B.Rmd",
                    "session_10C.Rmd"), 
                  text = paste("Download Session 10 R markdown files"), 
                  name = paste0("10_session", ".zip")
                  )
```

## Resources

### Effect sizes in general

https://www.statisticssolutions.com/statistical-analyses-effect-size/

https://www.scribbr.com/statistics/effect-size/

### Cohen's *d*

https://rpsychologist.com/cohend/

## Homework
The homework this week will be about statistical inference based on the Cohen's *d* effect size. First, write a function to calculate Cohen's *d*, taking sample size into account when calculating the pooled standard deviation. Next, try out your function on data of your choosing, and assess the probability that you would observe the Cohen's *d* (or something more extreme) you see in your data under the null hypothesis that the true *d* value in the population is zero. In other words, use a randomization test to calculate the p-value for the effect size. 

