---
title: "Cohen's *d*"
output: html_document
---

\

## Load packages and generate data
```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(MASS)
library(cowplot)

d1 <- tibble(values = c(mvrnorm(2000, 2, 1),
                        mvrnorm(2000, 2.5, 1)),
             group = rep(c("Ctrl", "Drug"), each = 2000))

d2 <- tibble(values = c(mvrnorm(2000, 2, 2^2),
                        mvrnorm(2000, 2.5, 2^2)),
             group = rep(c("Ctrl", "Drug"), each = 2000))

d3 <- tibble(values = c(mvrnorm(2000, 2, 4^2),
                        mvrnorm(2000, 2.5, 4^2)),
             group = rep(c("Ctrl", "Drug"), each = 2000))
```

\

## Cohen's *d*

The equation to calculate Cohen's *d* is shown below. The concept is quite simple mathematically; the mean value difference is divided by a combined measure of standard deviation often referred to as the pooled standard deviation. 

\

$$ 
Cohen's \,d = \frac{M_{2}-M_{1}}{pooled\,SD}
$$
\

There are different ways the pooled standard deviation can be calculated. Below is the simplest way of doing it, but this method can result in sub-optimal results if group sizes are very different.

\

$$
pooled\, SD = \sqrt{\frac{SD_{1}^2+ SD_{2}^2}{2}}
$$

\

A "safer" way of calculating the pooled standard deviation includes sample sizes, as shown below. 

\

$$
pooled\, SD = \sqrt{\frac{SD_{1}^2(n_{1}-1)+ SD_{2}^2(n_{2}-1)}{n_{1}+n_{2}-2}}
$$

\

where $SD_{1}$ is the standard deviation of group one, $SD_{2}$ is the standard deviation of group two, and $n_{1}$ and $n_{2}$ are the sample sizes of group one and two, respectively. 

\

## Function to calculate Cohen's *d*

```{r}
CD <- function(data, values, group){
  m_diff <- data %>% 
    group_by({{group}}) %>% 
    summarise(m = mean({{values}})) %>% 
    summarize(d = diff(m)) %>% 
    pull()
  
  p_sdev <- data %>% 
    group_by({{group}}) %>% 
    summarise(sd = sd({{values}})) %>% 
    transmute(sd_sq = sd^2) %>% 
    summarise(sqrt(sum(sd_sq)/2)) %>% 
    pull()
  
  return(m_diff/p_sdev)
} 
```

\

## Cohen's *d* results
```{r, message=FALSE}
CD(d1, values, group)
CD(d2, values, group)
CD(d3, values, group)

```

\

## Using `summarise`
```{r, message=FALSE}
d1 %>% 
  summarise(cd = CD(., values, group))
```


